A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-06-15 tex/sec2-outline.tex
[ tex ] Add the pen-and-paper proof of bidirectional type synthesis
[ tex ] Add points to address; finalise the paper title
[ tex ] Added acmart files; separate compilation for TeXShop
[ tex ] Using the same color as by Dunfield et al.
[ tex ] Fix some figures; add stubs for typesetting rules

%! TEX root = BiSig.tex

\section{Core Ideas with Bidirectional Simply Typed \texorpdfstring{$\lambda$}{lambda}-Calculus} \label{sec:key-ideas}
\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Gamma \vdash \isTerm{t} \Rightarrow A}
    \quad\text{and}\quad \boxed{\Gamma \vdash \isTerm{t} \Leftarrow A}
    \\
    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash \isTerm{x} \Rightarrow A}\;(\text{var}^\Rightarrow)
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \Leftarrow B \\ A = B}{\Gamma \vdash \isTerm{t : B}\Rightarrow A}\;(\text{anno})
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow B \\ A = B}{\Gamma \vdash \isTerm{t\subsum} \Leftarrow A}\;(\text{sub})
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow A \bto B \\ \Gamma \vdash \isTerm{u} \Leftarrow A}{\Gamma \vdash \isTerm{t\;u} \Rightarrow B}\;(\text{app})
    \and
    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} \Leftarrow B}{\Gamma \vdash \isTerm{\lam{x}t} \Leftarrow A \bto B}\;(\text{abs})
  \end{mathpar}
  \caption{Bidirectional simply typed $\lambda$-calculus $\Lambda^{\Leftrightarrow}_{\bto}$}
  \label{fig:bi-stlc}
\end{figure}
\begin{remark}
  We avoid using the term `function type' and its conventional notation $\to$ at the object level on purpose, as it may be confused with function types in our type-theoretic meta-language.
\end{remark}

>>>>

%!TEX root = BiSig.tex

\section{The Key Ideas} \label{sec:key-ideas}
\begin{figure}
  \small
  \bgroup
  \renewcommand{\arraystretch}{1.5}
  \begin{tabular}{ r l }
    $\boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A}$ & A raw term $t$ synthesises a type $A$ under $\Gamma$ \\
    $\boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} \chk A}$ &A raw term $t$ checks against a type $A$ under $\Gamma$
  \end{tabular}
  \egroup
  \centering
  \begin{mathpar}
    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash \isTerm{x} \syn A}\,\SynRule{Var}
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \chk B \\ A = B}{\Gamma \vdash \isTerm{t : B}\syn A}\,\SynRule{Anno}
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \syn B \\ A = B}{\Gamma \vdash \isTerm{t} \chk A}\,\ChkRule{Sub}
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \syn A \bto B \\ \Gamma \vdash \isTerm{u} \chk A}{\Gamma \vdash \isTerm{t\;u} \syn B}\,\SynRule{App}
    \and
    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} \chk B}{\Gamma \vdash \isTerm{\lam{x}t} \chk A \bto B}\,\ChkRule{Abs}
  \end{mathpar}
  \caption{Bidirectional simply typed $\lambda$-calculus $\Lambda^{\Leftrightarrow}_{\bto}$}
  \label{fig:bi-stlc}
\end{figure}
\begin{remark}
  We avoid using the term `function type' and its conventional notation $\to$ at the object level on purpose, as it may be confused with function types in our type-theoretic meta-language.
\end{remark}

### 2023-06-15 tex/sec3-defs.tex
[ tex ] Minor
[ tex ] Fix some figures; add stubs for typesetting rules

\begin{figure}
  \begin{minipage}[b]{.55\textwidth}
    \centering
    \small
    \judgbox{\Theta|-_{\Sigma} A}{$A$ is well-formed with type variables in $\Theta$}
    \begin{mathpar}
      \inferrule{\Theta \ni X_i}{\Theta |-_\Sigma X_i} \and
    \inferrule{\Theta |-_{\Sigma} A_1 \\ \cdots \\ \Theta |-_{\Sigma} A_{\arity(o)}}{\Theta |-_{\Sigma} \tyOp_o(A_1, \ldots, A_{\arity(o)})}
    \end{mathpar}
    \caption{Type formation rules}
    \label{fig:simple-type}
  \end{minipage}
  \begin{minipage}[b]{.4\textwidth}
    \centering
    \small
    \judgbox{\Theta |-_{\Sigma} \Gamma}{}
    \begin{mathpar}
      \inferrule{ }{\Theta |-_{\Sigma} \cdot }\and
      \inferrule{\Theta |-_{\Sigma} A \\ \Theta|-_{\Sigma} \Gamma}{\Theta |-_{\Sigma} \Gamma, A}
    \end{mathpar}
    \caption{Context formation rules}
  \label{fig:simple-context}
  \end{minipage}
\end{figure}

@@

\begin{figure}
  \centering
  \small
  \judgbox{|-_{\Sigma, \Omega} \isTerm{t}}{$t$ is a raw term for the signature $(\Sigma, \Omega)$}
  \begin{mathpar}
    \inferrule{x : \Identifier}{|-_{\Sigma, \Omega} \isTerm{x}}\,\Rule{Var}
    \and
    \inferrule{\cdot |-_{\Sigma} A \\ |-_{\Sigma, \Omega}\isTerm{t}}{|-_{\Sigma, \Omega} \isTerm{t \annote A}}\,\Rule{Anno}
    \\
    \inferrule{|-_{\Sigma, \Omega} \isTerm{t_1} \quad \cdots \quad |-_{\Sigma, \Omega} \isTerm{t_n}}
    {|-_{\Sigma, \Omega} \isTerm{\tmOp_o(t_1, \ldots, t_n)}}\,\Rule{Op}_o
    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n} \to A$ in $\Omega$}
  \end{mathpar}
  \caption{Raw term formation rules}
\end{figure}

@@

\begin{figure}
  \centering
  \small
  \judgbox{|-_{\Sigma, \Omega} \isTerm{t}^\dir{d}}{$t$ is a raw term for $(\Sigma, \erase{\Omega})$ in mode $d$ for  $(\Sigma, \Omega)$}
  \begin{mathpar}
    \inferrule{x : \Identifier}{|-_{\Sigma, \Omega} \isTerm{x}^{\syn}}\,\SynRule{Var}
    \and
    \inferrule{\cdot |-_{\Sigma} A \\ |-_{\Sigma, \Omega}\isTerm{t}^{\chk}}{|-_{\Sigma, \Omega} (\isTerm{t \annote A})^{\syn}}\,\SynRule{Anno}
    \and
    \inferrule{|-_{\Sigma, \Omega} \isTerm{t}^{\syn}}{|-_{\Sigma, \Omega} \isTerm{t}^{\chk}}\,\ChkRule{Sub}
  \end{mathpar}
  \begin{mathpar}
    \inferrule{|-_{\Sigma, \Omega} \isTerm{t_1}^\dir{d_1} \quad \cdots \quad |-_{\Sigma, \Omega} \isTerm{t_n}^\dir{d_n}}
    {|-_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\, t_1; \ldots;\vec{x}_n.\, t_n)}^\dir{d}}\,\Rule{Op}
    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A_{n}^{\dir{d_n}} \to A^{\dir{d}}$ in $\Omega$}
  \end{mathpar}
  \caption{Preprocessed raw terms}
\end{figure}

\begin{figure}
  \centering
  \small
  \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^\dir{d} A}{A raw term $t$ has a type $A$ (without type variables) under $\Gamma$}
  \begin{mathpar}
    \inferrule{(x : A) \in \Gamma}{\Gamma |-_{\Sigma, \Omega} \isTerm{x} :^{\syn} A}\,\SynRule{Var}
    \and
    \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}{\Gamma |-_{\Sigma, \Omega} (\isTerm{t \annote B}) :^{\syn}  A}\,\SynRule{Anno}
    \and
    \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{{\syn}} B \\ A = B}{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}\,\ChkRule{Sub}
    \\
    \inferrule{\rho\colon \Sub{\Xi}{\emptyset} \\ \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_{1}} :^\dir{d_1} \simsub{A_1}{\rho} \\
      \cdots \\
    \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} :^{\dir{d_n}} \simsub{A_{n}}{\rho}}
    {\Gamma |-_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} :^{\dir{d}} \simsub{A}{\rho}} \,\mathsf{Op}
    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A_{n}^{\dir{d_n}} \to A^\dir{d}$ in $\Omega$}
  \end{mathpar}
  \begin{minipage}{.15\linewidth}
  Abbreviations: 
  \end{minipage}
  \begin{minipage}{.8\linewidth}
    \begin{flushleft}
    $\boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A} \defeq \boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\syn} A}$ \quad A raw term $t$ synthesises a type $A$ under $\Gamma$ \\
    $\boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} \chk A} \defeq \boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}$ \quad A raw term $t$ checks against a type $A$ under $\Gamma$
    \end{flushleft}
  \end{minipage}
  \caption{Bidirectional typing rules}
\end{figure}

\begin{example}

\end{example}

>>>>

\begin{figure}
  \begin{minipage}[b]{.6\textwidth}
    \centering
    \small
    \judgbox{\Theta|-_{\Sigma} A}{$A$ is a well-formed type with type variables in $\Theta$}
    \begin{mathpar}
      \inferrule{\Theta \ni X_i}{\Theta |-_\Sigma X_i} \and
    \inferrule{\Theta |-_{\Sigma} A_1 \\ \cdots \\ \Theta |-_{\Sigma} A_{\arity(o)}}{\Theta |-_{\Sigma} \tyOp_o(A_1, \ldots, A_{\arity(o)})}
    \end{mathpar}
    \caption{Simple types}
    \label{fig:simple-type}
  \end{minipage}
  \begin{minipage}[b]{.35\textwidth}
    \centering
    \small
    \judgbox{\Theta |-_{\Sigma} \Gamma}{}
    \begin{mathpar}
      \inferrule{ }{\Theta |-_{\Sigma} \cdot }\and
      \inferrule{\Theta |-_{\Sigma} A \\ \Theta|-_{\Sigma} \Gamma}{\Theta |-_{\Sigma} \Gamma, A}
    \end{mathpar}
    \caption{Contexts}
  \label{fig:simple-context}
  \end{minipage}
\end{figure}

@@

\begin{figure}
  \centering
  \small
  \judgbox{|-_{\Sigma, \Omega} \isTerm{t}}{$t$ is a raw term for the signature $(\Sigma, \Omega)$}
  \begin{mathpar}
    \inferrule{x : \Identifier}{|-_{\Sigma, \Omega} \isTerm{x}}\,\Rule{Var}
    \and
    \inferrule{\cdot |-_{\Sigma} A \\ |-_{\Sigma, \Omega}\isTerm{t}}{|-_{\Sigma, \Omega} \isTerm{t \annote A}}\,\Rule{Anno}
    \\
    \inferrule{|-_{\Sigma, \Omega} \isTerm{t_1} \quad \cdots \quad |-_{\Sigma, \Omega} \isTerm{t_n}}
    {|-_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.t_1; \ldots; \vec{x}_n.t_n)}}\,\Rule{Op}
    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n} \to A$ in $\Omega$}
  \end{mathpar}
  \caption{Raw terms}
\end{figure}

@@

\begin{figure}
  \centering
  \small
  \judgbox{|-_{\Sigma, \Omega} \isTerm{t}^\dir{d}}{$t$ is a raw term for $(\Sigma, \erase{\Omega})$ in mode $d$ for  $(\Sigma, \Omega)$}
  \begin{mathpar}
    \inferrule{x : \Identifier}{|-_{\Sigma, \Omega} \isTerm{x}^{\syn}}\,\SynRule{Var}
    \and
    \inferrule{\cdot |-_{\Sigma} A \\ |-_{\Sigma, \Omega}\isTerm{t}^{\chk}}{|-_{\Sigma, \Omega} (\isTerm{t \annote A})^{\syn}}\,\SynRule{Anno}
    \and
    \inferrule{|-_{\Sigma, \Omega} \isTerm{t}^{\syn}}{|-_{\Sigma, \Omega} \isTerm{t}^{\chk}}\,\ChkRule{Sub}
  \end{mathpar}
  \begin{mathpar}
    \inferrule{|-_{\Sigma, \Omega} \isTerm{t_1}^\dir{d_1} \quad \cdots \quad |-_{\Sigma, \Omega} \isTerm{t_n}^\dir{d_n}}
    {|-_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\, t_1; \ldots;\vec{x}_n.\, t_n)}^\dir{d}}\,\Rule{Op}
    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A_{n}^{\dir{d_n}} \to A^{\dir{d}}$ in $\Omega$}
  \end{mathpar}
  \caption{Preprocessed raw terms, I}
\end{figure}

\begin{figure}
  \centering
  \small
  \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^\dir{d} A}{A raw term $t$ is of type $A$ (without type variables) in mode $d$ under $\Gamma$}
  \begin{mathpar}
    \inferrule{(x : A) \in \Gamma}{\Gamma |-_{\Sigma, \Omega} \isTerm{x} :^{\syn} A}\,\SynRule{Var}
    \and
    \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}{\Gamma |-_{\Sigma, \Omega} (\isTerm{t \annote B}) :^{\syn}  A}\,\SynRule{Anno}
    \and
    \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{{\syn}} B \\ A = B}{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}\,\ChkRule{Sub}
    \\
    \inferrule{\rho\colon \Sub{\Xi}{\emptyset} \\ \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_{1}} :^\dir{d_1} \simsub{A_1}{\rho} \\
      \cdots \\
    \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} :^{\dir{d_n}} \simsub{A_{n}}{\rho}}
    {\Gamma |-_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} :^{\dir{d}} \simsub{A}{\rho}} \,\mathsf{Op}
    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A_{n}^{\dir{d_n}} \to A^\dir{d}$ in $\Omega$}
  \end{mathpar}
Abbreviations: $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \syn A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^{\syn}}$ and $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \chk A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^{\chk}}$
  \caption{Bidirectional typing rules}
\end{figure}

\LT{What induction principle are we using?}
\begin{theorem}[Mutual Induction Principles]

\end{theorem}

### 2023-06-15 tex/sec4-bidirectionalisation.tex
[ tex ] Fix some figures; add stubs for typesetting rules

\subsection{Bidirectionalisation}
\begin{proposition}[Decidability for Bidirectionalisation] \label{thm:bidirectionalisation}
  It is decidable whether a raw term~$t$ is in mode~$d$ for any $t$ and $d$.
\end{proposition}

@@

>>>>

\subsection{Bidirectionalisation}

\begin{proposition}[Decidability for Bidirectionalisation] \label{thm:bidirectionalisation}
  It is decidable whether a raw term~$t$ is in mode~$d$ for any $t$ and $d$.
\end{proposition}

@@

\begin{figure}
  \centering
  \small
  \bgroup
  \renewcommand{\arraystretch}{1.5}
  \begin{tabular}{ r l }
    $\boxed{|-_{\Sigma, \Omega} \isTerm{t}\;\dir{d}}$ & $t$ is a raw term possibly with missing annotation or redundant casts in mode $d$ \\
    $\boxed{|-_{\Sigma, \Omega} \isTerm{t}\;\dir{d}}$ & $t$ is a raw term without missing annotation \\
    $\boxed{|-_{\Sigma, \Omega} \isTerm{t}\;\dir{d}}$ & $t$ is a raw term without redundant casts \\
  \end{tabular}
  \egroup
  \LT{Design a notation for \textsf{Pre?} so that a raw term without missing annotation or redundant casts is exactly the intersection of these two cases.}
  \begin{mathpar}
    \inferrule{x : \Identifier}{|-_{\Sigma, \Omega} \isTerm{x}^{\syn}}\,\SynRule{Var}
    \and
    \inferrule{|-_{\Sigma, \Omega}\isTerm{t}^{\chk}}{|-_{\Sigma, \Omega} (\isTerm{t \annote A})^{\syn}}\,\SynRule{Anno}
    \and
    \inferrule{|-_{\Sigma, \Omega}\isTerm{t}^{\chk}}{|-_{\Sigma, \Omega} \isTerm{t}^{\syn}}\,\SynRule{Anno?}
    \and
    \inferrule{|-_{\Sigma, \Omega} \isTerm{t}^{\syn}}{|-_{\Sigma, \Omega} \isTerm{t}^{\chk}}\,\ChkRule{Sub}
    \and
    \inferrule{|-_{\Sigma, \Omega} \isTerm{t_1}^\dir{d_1} \quad \cdots \quad |-_{\Sigma, \Omega} \isTerm{t_n}^\dir{d_n}}
    {|-_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\, t_1; \ldots;\vec{x}_n.\, t_n)}^\dir{d}}\,\Rule{Op}
  \end{mathpar}
  \caption{Preprocessed raw terms, II}
  \label{fig:raw-with-missing-annotations}
\end{figure}

\begin{figure}
  \centering\small
  \judgbox{\isTerm{t} \sqsupseteq \isTerm{u}}{A raw term $t$ is more annotated than $u$ (for some bidirectional type system $(\Sigma, \Omega)$)}
  \begin{mathpar}
    \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}{(\isTerm{t} : A) \sqsupseteq \isTerm{u}}\;\Rule{More}
    \and
    \inferrule{\vphantom{x : \Identifier}}{\isTerm{x} \sqsupseteq \isTerm{x}}
    \and
    \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}{(\isTerm{t} : A) \sqsupseteq (\isTerm{u} : A)}
    \and
    \inferrule{\isTerm{t_1} \sqsupseteq \isTerm{u_1} \quad \cdots \quad \isTerm{t_n} \sqsupseteq{u_n}}
    {\isTerm{\tmOp_o(\vec{x}_1. t_1; \ldots; \vec{x}_n. t_n)} \sqsupseteq \isTerm{\tmOp_o(\vec{x}_1.u_1; \ldots; \vec{x}_n.u_n)}}
  \end{mathpar}

  \caption{Annotation ordering between raw terms}
  \label{fig:annotation-order}
\end{figure}
